Analytical Predictions of Explanet Obliquities Generated by Planet-Disk Interactions

### 2020-01-22 writeups/Su_disk_disp.tex
stash DL first half comments

\section{Introduction}

@@

We define the frequency ratio $\eta$ and dimensionless time $\tau$ as
\begin{align}
    \eta \equiv{}& -\frac{g}{\alpha}\label{eq:eta},\\
        ={}& 3.75 \p{\frac{k_{\rm p}}{6k_{\rm qp}}}
            \p{\frac{m_{\rm p}}{0.001 M_\odot}}
            \p{\frac{R_{\rm p}}{R_J}}^{-3}
        &\p{\frac{M_{\rm d}}{0.01 M_{\odot}}}
            \p{\frac{a_{\rm p}}{1\;\mathrm{AU}}}^{9/2}
            \p{\frac{r_{\rm d}}{30 \;\mathrm{AU}}}^{-3}
            \p{\frac{M_\star}{1M_{\odot}}}^{-3/2}
            \cos I,\\
    \tau \equiv{}& \alpha t.
\end{align}
Then Eq.~\eqref{eq:dsdt_rot} becomes
\begin{equation}
    \at{\rd{\hat{s}}{t}}_{\rm rot} = \p{\hat{s} \cdot \hat{l}}
            \p{\hat{s} \times \hat{l}}
        - \eta\p{\hat{s} \times \hat{l}_{\rm d}}. \label{eq:dsdt_base}
\end{equation}

Throughout this paper, we consider $\alpha$ constant, but allow $g$ to vary in
time. In the dispersing disk scenario of \citet{millholland_disk}, $\abs{g}$
decreases in time due to the decreasing disk mass. We consider a simple
exponential decay model
\begin{equation}
    M_{\rm d}(t) = M_{\rm d}(0)e^{-t/t_{\rm d}},
\end{equation}
with $t_d$ constant.

Tihs implies $\rd{\eta}{t} = -\frac{\eta}{t_{\rm d}}$. In dimenisonless units,
we have
\begin{equation}
    \rd{\eta}{\tau} = -\epsilon \eta\label{eq:deta_dt},
\end{equation}
where
\begin{align}
    \epsilon \equiv{}& \frac{1}{\alpha t_{\rm d}},\\
        ={}& 0.015 \p{\frac{k_{\rm p}}{6k_{\rm qp}}}
            \p{\frac{m_{\rm p}/M_\star}{0.001}}
            \p{\frac{a_{\rm p}}{1\;\mathrm{AU}}}^3\nonumber\\
        &\p{\frac{R_{\rm p}}{R_J}}^{-3}
            \p{\frac{\Omega_{\rm p}}{2\pi/\p{10\;\mathrm{days}}}}
            \p{\frac{t_{\rm d}}{10\;\mathrm{Myr}}}^{-1}.
\end{align}
Eqs.~\eqref{eq:dsdt_base} and~\eqref{eq:deta_dt} together constitute our system
of study.

@@

Spin states satisfying $\at{\rd{\hat{s}}{t}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs). They require that $\hat{s}$, $\hat{l}$, and
$\hat{l}_{\rm d}$ be coplanar. There are either two or four solutions, depending
on the values of $\eta, I$. They are specified by the obliquity $\theta$ and the
precessional phase of $\hat{s}$ around $\hat{l}$, denoted $\phi$. Following
standard convention and nomenclature (see Fig.~\ref{fig:cs_locs}), CSs 1, 3, 4
have $\phi = 0$ and $\theta < 0$, corresponding to $\hat{s}$ and $\hat{l}_{\rm
d}$ being on opposite sides of $\hat{l}$, while CS2 has $\phi = \pi$ and $\theta
> 0$, corresponding to $\hat{s}$ and $\hat{l}_{\rm d}$ being on the same side of
$\hat{l}$. When $\eta < \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini state obliquities as a function of $\eta$ at $I = 5^\circ$.
    Note that $\theta \in [-\pi, \pi]$ is the traditional definition of the
    polar angle. The vertical dashed line indicates $\eta_{\rm c}$, where CS1
    and CS4 merge and annihilate, and the horizontal dashed lines indicate
    $\theta = I$ and $180^\circ - I$, the asymptotic behavior for CSs 2 and 3
    for $\eta \gg \eta_{\rm c}$.}\label{fig:cs_locs}
\end{figure}

@@

The unsigned areas $\p{\abs{\int \cos \theta \;\mathrm{d}\phi}}$ of the three
zones can be computed analytically \citep{henrard1987,ward2004I}. Defining
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},\\
    \end{align}
\end{subequations}
the areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},\\
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2},\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_{\rm c}$ since the
separatrix disappears, but a natural extension exists: evolve an initial phase
space point $p$ under adiabatic decrease of $\eta$ until the separatrix appears
at $\eta = \eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm
c}$. Since phase space area is conserved under adiabatic evolution, this implies
$A_i\p{\eta > \eta_{\rm c}} = A_i(\eta_{\rm c})$. The boundary between these
extended zones is denoted by the dashed black line in the top left panel of
Fig.~\ref{fig:eq_1contours}, where otherwise no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/1_areas.png}
    \caption{Plot of fractional areas of each of the zones $A_{\rm i}(\eta) /
    4\pi$ as given by Eqs.~\eqref{se:area_ward}. Dotted lines correspond to
    small $\eta$ approximations discussed in Appendix~\ref{s:ad_approx}. Dashed
    lines for $\eta > \eta_{\rm c}$ effective values of $A_{\rm II}, A_{\rm
    III}$ for $\eta > \eta_{\rm c}$, denoting the area of points that would flow
    into either area under adiabatic decrease of $\eta$ from $\eta > \eta_{\rm
    c}$.}\label{fig:eq_areas}
\end{figure}

@@

If a Hamiltonian system is parameterized by some parameter that is varied, its
evolution is considered \emph{adiabatic} if the parameter's variation is slow
compared to the Hamiltonian dynamics. In the current problem, the period of the
longest orbit of the constant-$\eta$ Hamiltonian system is infinite, namely
along the separatrix. This is an essential feature of the problem, as otherwise
trajectories would never be able to experience separatrix crossings. But this
seemingly implies adiabaticity is impossible to define.

Instead of requiring the variation of $\eta$ be slower than all timescales in
the Hamiltonian dynamics, we seek a weaker but nevertheless effective criterion.
In practice, trajectories only cross the separatrix once, while they spend many
orbits inside one of the three zones and far from the separatrix as $\eta$
evolves. Thus, a \emph{weak adiabaticity criterion} is that the variation of
$\eta$ be slower than all circulation/libration timescales in the Hamiltonian
system. If this criterion is satisfied, then a given trajectory will be
adiabatic for almost all of its evolution save one or two separatrix-crossing
orbits.

As is shown in Appendix~\ref{ss:eigens}, the libration frequency about CS2 is slower
than that about CS1 or CS3. As such, it is the slowest characteristic frequency
in the system. The weak adiabaticity criterion can be recast as requiring that
$\eta$ vary more slowly than this libration period; this is exactly the
criterion used in \citep{ward2004II,millholland_disk}. This constraint can be
computed
\begin{align}
    -\at{\frac{1}{\eta}\rd{\eta}{\tau}} = \epsilon &\lesssim
            \frac{1}{T_{\rm lib}},\\
        &\lesssim \frac{1}{2\pi}\sqrt{\eta\sin I \sin \theta_2
            \p{1 + \eta \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{align}
where $\theta_2$ is the obliquity at CS2. This formula differs somewhat from
that presented in \citet{millholland_disk}, where the $\csc^3\theta_2$ term is
neglected and a square root is missing.

For $I = 5^\circ$, $\eta \sim \eta_c$ we compute $\epsilon \lesssim 0.085$ is
the requirement for weak adiabaticity. Since our criterion is only a weak
constraint on adiabaticity, we use $\epsilon = 3 \times 10^{-4}$ in our
``adiabatic'' simulations. We explore the consequences of increasingly
nonadiabatic evolution in Section~\ref{ss:transition}.

@@

We consider the evolution of a system with arbitrary initial $\theta_{\rm sd, i}
\equiv \at{\theta_{\rm sd}}_{t = 0}$ and $\eta_i \equiv \at{\eta}_{t = 0} \gg
1$. We are interested in the distribution of final obliquities after $\eta$
decreases to $\eta_f \ll 1$ (after the disk has dissipated to negligible mass).
While it is likely that $\theta_{\rm sd, i} \approx 0$ as the planet forms in
the disk, we explore all $\theta_{\rm sd, i} \in [0, \pi]$ for completeness.

We evenly sample $101$ values of $\theta_{\rm sd, i}$, and for each $\theta_{\rm
sd, i}$ value, we pick $101$ $\hat{s}$ initial conditions approximately on the
ring fixed by $\theta_{\rm sd, i}$\footnote{For each $\theta_{\rm sd, i}$ we
choose an initial condition on the libration cycle going through $\p{\theta_2 +
\theta_{\rm sd, i}, \phi_2}$ (where $\p{\theta_2, \phi_2}$ are the coordinates
of CS2), such that all initial conditions for a given $\theta_{\rm sd, i}$
enclose the same initial phase space; see Section~\ref{sss:a_evo}.}. To be concrete,
we choose $\eta_i = 10\eta_{\rm c}$. For each initial condition, we evolve
Equations~\ref{eq:dsdt_base} and~\ref{eq:deta_dt} until $\eta = 10^{-5}$. For
such small $\eta$, $\hat{s}$ precesses about $\hat{l}$ with uniform final
obliquity $\theta_{\rm f}$. This transfer function $\theta_{\rm f}\p{\theta_{\rm
sd, i}}$ is our primary result, and is shown for $I = 5^\circ, 10^\circ,
20^\circ$ in Figs.~\ref{fig:ad_ensemble},~\ref{fig:3_ensemble_10_35},
and~\ref{fig:3_ensemble_20_35} respectively. The blue dots represent the results
of the simulation, which are in excellent agreement with the colored tracks.
These colored tracks are calculated semi-analytically using methods explained in
the following section.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_05_35.png}
    \caption{Top: $\theta_{\rm f}\p{\theta_{\rm sd, i}}$, overlaid with
    semi-analytic predictions of the $\theta_{\rm  f}$ for each of the four
    nontrivial dynamical tracks in colored lines. Black dotted lines correspond
    to small-$\theta_{\rm sd, i}$ predictions derived in
    Appendix~\ref{s:ad_approx}. Bottom: Semi-analytic probabilities of each of
    the dynamical tracks for each $\theta_{\rm sd, i}$. The five regimes of
    $\theta_{\rm sd, i}$ values demarcated by the vertical dashed black lines
    correspond to the five regimes of $A_i$ calculated in
    Section~\ref{ss:zone_transitions}.}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_10_35.png}
    \caption{Same as Fig.~\ref{fig:ad_ensemble} but for $I =
    10^\circ$ and with fewer annotations.}\label{fig:3_ensemble_10_35}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_20_35.png}
    \caption{Same as Fig.~\ref{fig:3_ensemble_10_35} but for $I = 20^\circ$. The
    $III \to II \to I$ track is very short since the maximum $A_{\rm II}$ is very
    near $\eta_{\rm c}$ for $I = 20^\circ$.}\label{fig:3_ensemble_20_35}
\end{figure}

\subsection{Adiabatic Evolution Theory}\label{ss:zone_transitions}

The evolutionary tracks that govern the behavior of $\theta_{\rm
f}\p{\theta_{\rm sd, i}}$ correspond to particular sequences of separatrix
crossings. Coarsely, they can be understood using a combination of (i) how the
enclosed phase space area by the trajectory evolves across each separatrix
crossing, and (ii) the associated probabilities with each separatrix crossing.

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. In
particular, we adopt convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
This definition of $A$ has the advantage of (i) being continuous at $\cos \theta
= 1$, and (ii) being easily expressible as combinations of the $A_i$ of
Eqs.~\eqref{se:area_ward}. The bounds of the integral are either a libration
($\phi$ returns to its original value with the same $\dot{q\phi}$ sign) or
circulation cycle ($\phi$ advances to $\phi \pm 2\pi$). Note lastly that $A$ has
simple physical interpretation of are enclosed on the unit sphere by $\hat{s}$
over one cycle measured relative to $+\hat{z}$:
\begin{equation}
    \int \int\limits_{\cos \theta(\phi)}^1
        \mathrm{d}\cos\theta\;\mathrm{d}\phi = \int \p{1 - \cos \theta(\phi)}
            \;\mathrm{d}\phi.
\end{equation}

Observe that when $\eta_i \gg 1$, trajectories librate about $\hat{l}_{\rm d}$
with constant $\theta_{\rm sd}$, meaning they enclose initial phase space area
\begin{equation}
    A_i \equiv \at{A}_{t = 0}
        = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_i$, as trajectories near CS2
or CS3 librate about these equilibria respectively, rather than $\hat{l}_{\rm
d}$, and Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice,
Eq.~\eqref{eq:ai_qsd} holds very well when defining $\theta_{\rm sd, i}$ as the
angular distance to CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

Beginning at the last separatrix crossing, the final enclosed phase space area
$A_f$ will be conserved for all time. As $\eta \to 0$, trajectories circulate at
constant obliquity $\theta_{\rm f}$, related to $A_f$ by
\begin{equation}
    2\pi\p{1 - \cos \theta_{\rm f}} = A_f. \label{eq:qfaf}
\end{equation}

@@

When a trajectory experiences separatrix crossing, it transitions into nearby
zones probabilistically. This process is well understood by
\citealp{henrard1982,henrard1987}. Their results may be simply summarized as
follows: if a zone $i$ is shrinking while adjacent zones $j, k$ are expanding
such that the sum of their areas is constant, the probability of transition from
zone $i$ to zone $j$ is given
\begin{align}
    \Pr\p{i \to j} = -\frac{\pd{A_j}{t}}{ \pd{A_i}{t}}
        = -\frac{\pd{A_j}{\eta}}{ \pd{A_i}{\eta}},\\
    \Pr\p{i \to k}
        = -\frac{\pd{A_k}{\eta}}{ \pd{A_i}{\eta}}.\label{eq:henrard_hop}
\end{align}
Note that $\Pr \p{i \to j} + \Pr\p{i \to k} = 1$. This can be used directly in
conjunction with Eqs.~\eqref{se:area_ward} to understand for what initial
conditions each track can be observed and with what probabilities.

@@

It should be noted that this result breaks down in two ways:
\begin{enumerate}
    \item If the evolution of $\hat{s}$ is adiabatic, then it is invalid to
        assume $s_z$ is approximately constant in time, as many
        circulation/libration orbits can ensue. Only when the driving is
        sufficiently impulsive that the evolution dominates the change in $s_z$
        is this calculation valid.

>>>>

\section{Introduction}\label{s:intro}

@@

We define the frequency ratio $\eta$ and dimensionless time $\tau$ as
\begin{align}
    \eta \equiv{}& -\frac{g}{\alpha}\label{eq:eta},\\
        ={}& 3.75 \p{\frac{k_{\rm p}}{6k_{\rm qp}}}
            \p{\frac{m_{\rm p}}{0.001 M_\odot}}
            \p{\frac{R_{\rm p}}{R_{\rm J}}}^{-3}
        &\p{\frac{M_{\rm d}}{0.01 M_{\odot}}}
            \p{\frac{a_{\rm p}}{1\;\mathrm{AU}}}^{9/2}
            \p{\frac{r_{\rm d}}{30 \;\mathrm{AU}}}^{-3}
            \p{\frac{M_\star}{1M_{\odot}}}^{-3/2}
            \cos I,\\
    \tau \equiv{}& \alpha t.
\end{align}
Then Eq.~\eqref{eq:dsdt_rot} becomes
\begin{equation}
    \at{\rd{\hat{s}}{t}}_{\rm rot} = \p{\hat{s} \cdot \hat{l}}
            \p{\hat{s} \times \hat{l}}
        - \eta\p{\hat{s} \times \hat{l}_{\rm d}}. \label{eq:dsdt_base}
\end{equation}

Throughout this paper, we consider $\alpha$ constant, but allow $g$ to vary in
time. In the dispersing disk scenario of \citet{millholland_disk}, $\abs{g}$
decreases in time due to the decreasing disk mass. We consider a simple
exponential decay model
\begin{equation}
    M_{\rm d}(t) = M_{\rm d}(0)e^{-t/t_{\rm d}},\label{eq:dmd_dt}
\end{equation}
with $t_{\rm d}$ constant.

Tihs implies $\rd{\eta}{t} = -\frac{\eta}{t_{\rm d}}$. In dimenisonless units,
we have
\begin{equation}
    \rd{\eta}{\tau} = -\epsilon \eta\label{eq:deta_dt},
\end{equation}
where
\begin{align}
    \epsilon \equiv{}& \frac{1}{\alpha t_{\rm d}},\label{eq:eps_def}\\
        ={}& 0.015 \p{\frac{k_{\rm p}}{6k_{\rm qp}}}
            \p{\frac{m_{\rm p}/M_\star}{0.001}}
            \p{\frac{a_{\rm p}}{1\;\mathrm{AU}}}^3\nonumber\\
        &\p{\frac{R_{\rm p}}{R_{\rm J}}}^{-3}
            \p{\frac{\Omega_{\rm p}}{2\pi/\p{10\;\mathrm{days}}}}
            \p{\frac{t_{\rm d}}{10\;\mathrm{Myr}}}^{-1}.
\end{align}
Eqs.~\eqref{eq:dsdt_base} and~\eqref{eq:deta_dt} together constitute our system
of study.

@@

Spin states satisfying $\at{\rd{\hat{s}}{t}}_{\rm rot} = 0$ are referred to as
\emph{Cassini States} (CSs). They require that $\hat{s}$, $\hat{l}$, and
$\hat{l}_{\rm d}$ be coplanar. There are either two or four solutions, depending
on the values of $\eta, I$. They are specified by the obliquity $\theta$ and the
precessional phase of $\hat{s}$ around $\hat{l}$, denoted $\phi$. Following
standard convention and nomenclature (see Fig.~\ref{fig:cs_locs}), CSs 1, 3, 4
have $\phi = 0$ and $\theta < 0$, corresponding to $\hat{s}$ and $\hat{l}_{\rm
d}$ being on opposite sides of $\hat{l}$, while CS2 has $\phi = \pi$ and $\theta
> 0$, corresponding to $\hat{s}$ and $\hat{l}_{\rm d}$ being on the same side of
$\hat{l}$. When $\eta < \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}I + \cos^{2/3}I}^{-3/2},\label{eq:etac}
\end{equation}
all four CSs exist, and when $\eta > \eta_{\rm c}$, only CSs 2, 3 exist. The CS
obliquities as a function of $\eta$ are shown in Fig.~\ref{fig:cs_locs}.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini state obliquities as a function of $\eta$ at $I = 5^\circ$.
    Note that $\theta \in [-\pi, \pi]$ is the traditional definition of the
    polar angle. The vertical dashed line indicates $\eta_{\rm c}$, where CS1
    and CS4 merge and annihilate, and the horizontal dashed lines indicate
    $\theta = I$ and $180^\circ - I$, the asymptotic behavior for CSs 2 and 3
    for $\eta \gg \eta_{\rm c}$.}\label{fig:cs_locs}
\end{figure}

@@

The unsigned areas $\p{\abs{\int \cos \theta \;\mathrm{d}\phi}}$ of the three
zones can be computed analytically \citep{henrard1987,ward2004I}. Defining
\begin{subequations}
    \begin{align}
        z_0 &= \eta\cos I, &
        \chi &= \sqrt{-\frac{\tan^3\theta_4}{\tan I} - 1},\\
        \rho &= \chi \frac{\sin^2 \theta_4\cos \theta_4}{
            \chi^2 \cos^2\theta_4 + 1},&
        T &= 2\chi \frac{\cos \theta_4}{
            \chi^2 \cos^2\theta_4 - 1},\\
    \end{align}
\end{subequations}
the areas for $\eta < \eta_{\rm c}$ are given by
\begin{subequations}\label{se:area_ward}
    \begin{align}
        A_{\rm II} &= 8\rho + 4\arctan T - 8z_0 \arctan \frac{1}{\chi},\\
        A_{\rm I} &= 2\pi\p{1 - z_0} - \frac{A_2}{2},\\
        A_{\rm III} &= 2\pi\p{1 + z_0} - \frac{A_2}{2}.
    \end{align}
\end{subequations}
These are plotted as a function of $\eta$ in Fig.~\ref{fig:eq_areas}. Note that
the zones are not formally defined for $\eta > \eta_{\rm c}$ since the
separatrix disappears, but a natural extension exists: evolve an initial phase
space point $p$ under adiabatic decrease of $\eta$ until the separatrix appears
at $\eta = \eta_{\rm c}$, then identify $p$ with the zone it is in at $\eta_{\rm
c}$. Since phase space area is conserved under adiabatic evolution, this implies
$A_{\rm i}\p{\eta > \eta_{\rm c}} = A_{\rm i}(\eta_{\rm c})$. The boundary between these
extended zones is denoted by the dashed black line in the top left panel of
Fig.~\ref{fig:eq_1contours}, where otherwise no separatrix exists.
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/99_misc/1_areas.png}
    \caption{Plot of fractional areas of each of the zones $A_{\rm i}(\eta) /
    4\pi$ as given by Eqs.~\eqref{se:area_ward}. Dotted lines correspond to
    small $\eta$ approximations discussed in Appendix~\ref{s:ad_approx}. Dashed
    lines for $\eta > \eta_{\rm c}$ effective values of $A_{\rm II}, A_{\rm
    III}$ for $\eta > \eta_{\rm c}$, denoting the area of points that would flow
    into either area under adiabatic decrease of $\eta$ from $\eta > \eta_{\rm
    c}$.}\label{fig:eq_areas}
\end{figure}

@@

In this section, we study the evolution of the planetary obliquity $\theta$ when
the parameter $\eta$ (or the disk mass $M_{\rm d}$; see Eq.~\eqref{eq:eta})
decreases sufficiently slowly that the evolution is adiabatic. Strictly
speaking, this requires the disk evolution timescale $t_{\rm d}$
(Eq.~\eqref{eq:dmd_dt}) be much larger than all timesccales of the dynamical
system governed by the Hamiltonian~\eqref{eq:H}. This is of course not possible
in all cases, as the motion along the separatrix has an infinite period.

In practice, as $\eta$ evolves, the system only crosses the separatrix once or
twice, while it spends many orbits inside one of the three zones and far from
the separatrix. Thus, a \emph{weak adiabaticity criterion} is that the variation
timescale of $\eta$ be slower than all circulation/libration timescales near the
equilibria/fixed points. If this criterion is satisfied, then the system will
evolve adiabatically for most of its evolution save one or two separatrix
crossings.

As is shown in Appendix~\ref{ss:eigens}, libration about CS2 is slower than that
about CS1 or CS3. As such, it is the slowest characteristic frequency in the
system. The weak adiabaticity criterion can be recast as requiring that $\eta$
vary more slowly than this libration frequency
\citep[see][]{ward2004II,millholland_disk}, i.e.\ $t_{\rm d}^{-1} \lesssim
\omega_{\rm lib} / 2\pi$. In terms of the dimensionless parameter $\epsilon$
(Eq.~\eqref{eq:eps_def}), this adiabaticity condition is
\begin{equation}
    \epsilon = \frac{1}{\alpha t_{\rm d}}
        \lesssim \frac{1}{2\pi}\sqrt{\eta\sin I \sin \theta_2
            \p{1 + \eta \sin I \csc^3 \theta_2}},
            \label{eq:ad_constr}
\end{equation}
where $\theta_2$ is the obliquity at CS2. This formula differs somewhat from
that presented in \citet{millholland_disk}, where the $\csc^3\theta_2$ term is
neglected and a square root is missing.

For $I = 5^\circ$, $\eta \sim \eta_{\rm c}$ Eq.~\eqref{eq:ad_constr} gives $\epsilon
\lesssim 0.085$. Since our criterion is only a weak condition for adiabaticity,
we use $\epsilon = 3 \times 10^{-4}$ in our ``adiabatic'' simulations below
(Section~\autoref{ss:ad_ensemble}). We explore the consequences of nonadiabatic
evolution in Section~\ref{s:nonad}.

@@

We consider the evolution of a system with arbitrary initial spin-disk
misalignment angle $\theta_{\rm sd, i}$ and initial $\eta_{\rm i} \gg 1$. We are
interested in the final spin obliquities $\theta_{\rm f}$ after $\eta$ gradually
decreases to $\eta_{\rm f} \ll 1$ (i.e.\ after the disk has dissipated to a
negligible mass). Note that when $\eta_{\rm i} \gg 1$, $\hat{l}$ precesses
around $\hat{l}_{\rm d}$ much faster than the spin-orbit precession
($\abs{\omega_{\rm ld}} \gg \abs{\omega_{\rm sl}}$), and the spin obliquity
$\theta$ is not well defined. So it is more appropriate to use $\theta_{\rm sd}$
rather than $\theta$ to specify the initial spin orientation.

As discussed in Section~\ref{s:intro}, we explore $\theta_{\rm sd, i} \in [0,
\pi]$. We choose $\epsilon = 3 \times 10^{-4}$ so that the system evolves almost
adiabatically (see above).

To obtain the distribution of the final obliquities $\theta_{\rm f}$, we evenly
sample $101$ values of $\theta_{\rm sd, i}$, and for each $\theta_{\rm sd, i}$
value, we pick $101$ $\hat{s}$ initial conditions approximately on the ring
fixed by $\theta_{\rm sd, i}$%
%
\footnote{For each $\theta_{\rm sd, i}$ we choose an initial condition on the
libration cycle going through $\p{\theta_2 + \theta_{\rm sd, i}, \phi_2}$ (where
$\p{\theta_2, \phi_2}$ are the coordinates of CS2), such that all initial
conditions for a given $\theta_{\rm sd, i}$ enclose the same initial phase
space; see Section~\ref{sss:a_evo}.}.
%
To be concrete, we choose $\eta_{\rm i} = 10\eta_{\rm c}$ where $\eta_{\rm c}$
is given by Eq.~\eqref{eq:etac} and evolve Equations~\ref{eq:dsdt_base}
and~\ref{eq:deta_dt} until $\eta$ reaches its final value $10^{-5}$. At such a
small $\eta$, $\hat{s}$ is strongly coupled to $\hat{l}$ and the final obliquity
$\theta_{rm f}$ is frozen. The mapping between $\theta_{\rm sd, i}$ and
$\theta_{\rm f}$ is our primary result, and is shown for $I = 5^\circ, 10^\circ,
20^\circ$ in Figs.~\ref{fig:ad_ensemble},~\ref{fig:3_ensemble_10_35},
and~\ref{fig:3_ensemble_20_35} respectively. The blue dots represent the results
of the numerical calculation. The colored tracks are calculated
semi-analytically using the method discussed in the following subsection.

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_05_35.png}
    \caption{Top: The final spin obliquity $\theta_{\rm f}$ as a function of the
    initial spin-disk misalignment angle $\theta_{\rm sd, i}$ for systems
    evolving from initial $\eta_{\rm i} \gg 1$ to $\eta_{\rm i} \ll 1$. The dots
    are results of numerical calculations (Section~\ref{ss:ad_ensemble}), and
    the colored tracks are semi-analytical results
    (Section~\ref{ss:zone_transitions}). The black dotted lines represent
    analytical approximations valid in the small-$\theta_{\rm sd, i}$ limit (see
    Appendix~\ref{s:ad_approx}). Bottom: Where a particular $\theta_{\rm sd, i}$
    corresponds to multiple tracks, the system evolves probabilistically.
    The track that a particular system evolves along in a numerical simulation
    can be measured by examining its final obliquity. The dots represent the
    inferred probabilities from measured final obliquities in our simulations,
    while the colored tracks denote the semi-analytic probability of the system
    evolving along each track. There are five regimes of $\theta_{\rm sd, i}$
    values for which different tracks are accessible. Semi-analytical
    calculations of the boundaries of these regimes are denoted via the vertical
    dashed black lines (see
    Section~\ref{ss:zone_transitions}).}\label{fig:ad_ensemble}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_10_35.png}
    \caption{Same as Fig.~\ref{fig:ad_ensemble} but for $I =
    10^\circ$ and with fewer annotations.}\label{fig:3_ensemble_10_35}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{../initial/2_toy2/3_ensemble_20_35.png}
    \caption{Same as Fig.~\ref{fig:3_ensemble_10_35} but for $I = 20^\circ$.
    % Note that the $III \to II \to I$ track is very short since the maximum
    % $A_{\rm II}$ is very near $\eta_{\rm c}$ for $I = 20^\circ$. (not a
    % clear/useful observation - yubo)
    }\label{fig:3_ensemble_20_35}
\end{figure}

\subsection{Analytical Theory for Adiabatic Evolution}\label{ss:zone_transitions}

The evolutionary tracks that govern the $\theta_{\rm f}$-$\theta_{\rm sd, i}$
mapping correspond to various sequences of separatrix crossings. They can be
understood using the principle of adiabatic invariance, combined with (i) how
the enclosed phase space area by the trajectory evolves across each separatrix
crossing, and (ii) the associated probabilities with each separatrix crossing.

@@

First, we consider how the enclosed phase space area by a trajectory evolves
over time. In the absence of separatrix encounters, the enclosed phase space
area $\oint \cos\theta \;\mathrm{d}\phi$ is an adiabatic invariant. In
particular, we adopt convention where
\begin{equation}
    A \equiv \oint \p{1 - \cos \theta}\;\mathrm{d}\phi.\label{eq:a_oint}
\end{equation}
This definition of $A$ has the advantage of (i) being continuous at $\cos \theta
= 1$, and (ii) being easily expressible as combinations of the $A_{\rm i}$ of
Eqs.~\eqref{se:area_ward}. The bounds of the integral are either a libration
($\phi$ returns to its original value with the same $\dot{q\phi}$ sign) or
circulation cycle ($\phi$ advances to $\phi \pm 2\pi$). Note lastly that $A$ has
simple physical interpretation of are enclosed on the unit sphere by $\hat{s}$
over one cycle measured relative to $+\hat{z}$:
\begin{equation}
    \int \int\limits_{\cos \theta(\phi)}^1
        \mathrm{d}\cos\theta\;\mathrm{d}\phi = \int \p{1 - \cos \theta(\phi)}
            \;\mathrm{d}\phi.
\end{equation}

Observe that when $\eta_{\rm i} \gg 1$, trajectories librate about $\hat{l}_{\rm d}$
with constant $\theta_{\rm sd}$, meaning they enclose initial phase space area
\begin{equation}
    A_{\rm i} \equiv \at{A}_{t = 0}
        = 2\pi\p{1 - \cos \theta_{\rm sd, i}}.\label{eq:ai_qsd}
\end{equation}
Complications arise when considering finite $\eta_{\rm i}$, as trajectories near CS2
or CS3 librate about these equilibria respectively, rather than $\hat{l}_{\rm
d}$, and Eq.~\eqref{eq:ai_qsd} is no longer exact. In practice,
Eq.~\eqref{eq:ai_qsd} holds very well when defining $\theta_{\rm sd, i}$ as the
angular distance to CS2; an exception is discussed in Section~\ref{sss:evol_traj}.

Beginning at the last separatrix crossing, the final enclosed phase space area
$A_{\rm f}$ will be conserved for all time. As $\eta \to 0$, trajectories circulate at
constant obliquity $\theta_{\rm f}$, related to $A_{\rm f}$ by
\begin{equation}
    2\pi\p{1 - \cos \theta_{\rm f}} = A_{\rm f}. \label{eq:qfaf}
\end{equation}

@@

When a trajectory experiences separatrix crossing, it transitions into nearby
zones probabilistically. This process is well understood by
\citealp{henrard1982,henrard1987}. Their results may be simply summarized as
follows: if a zone $i$ is shrinking while adjacent zones $j, k$ are expanding
such that the sum of their areas is constant, the probability of transition from
zone $i$ to zone $j$ is given
\begin{align}
    \Pr\p{i \to j} = -\frac{\pd{A_{\rm j}}{t}}{ \pd{A_{\rm i}}{t}}
        = -\frac{\pd{A_{\rm j}}{\eta}}{ \pd{A_{\rm i}}{\eta}},\\
    \Pr\p{i \to k}
        = -\frac{\pd{A_{\rm k}}{\eta}}{ \pd{A_{\rm i}}{\eta}}.\label{eq:henrard_hop}
\end{align}
Note that $\Pr \p{i \to j} + \Pr\p{i \to k} = 1$. This can be used directly in
conjunction with Eqs.~\eqref{se:area_ward} to understand for what initial
conditions each track can be observed and with what probabilities.

@@

It should be noted that this result breaks down in two ways:
\begin{enumerate}
    \item If the evolution of $\hat{s}$ is adiabatic, then it is invalid to
        assume $s_{\rm z}$ is approximately constant in time, as many
        circulation/libration orbits can ensue. Only when the driving is
        sufficiently impulsive that the evolution dominates the change in
        $s_{\rm z}$
        is this calculation valid.
